---
title: "The BUGS language (state-space model example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```
```{r loadnimble, include=FALSE}
library(nimble)
```

# The BUGS language

The BUGS language is described in detail in the [relevant section of the BUGS manual](http://www.openbugs.net/Manuals/ModelSpecification.html). 

### Stochastic declarations:

  - `x ~ dgamma(shape, scale)` (x follows a gamma distribution with parameters `shape` and `scale`)

### Deterministic declarations

  - `y <- 2 * x`

### For loops

  - `for(i in 1:10) {x[i] <- exp(y[i])}`

### Classic BUGS examples

  - included in NIMBLE or found here on the [JAGS website](https://sourceforge.net/projects/mcmc-jags/files/Examples/).
  - Information: [Volume 1](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol1.pdf), [Volume 2](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol2.pdf), and [Volume 3](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol3.pdf) of the WinBUGS example manuals.

### NIMBLE's dialect of BUGS :

   - Differences between NIMBLE and WinBUGS, OpenBUGS and/or JAGS will be covered later.
   - Also described in Section 5 of the [NIMBLE Users Manual](http://r-nimble.org/manuals/NimbleUserManual.pdf).

# A basic example: population state-space model

We'll use an example from Chapter 5 of K&eacute;ry and Schaub (2012)[^1]

Code and data are available from the [BPA web site](http://www.vogelwarte.ch/de/projekte/publikationen/bpa/complete-code-and-data-files-of-the-book.html), in bpa-code.txt.

### Summary

   - House martin counts (1990 - 2009) by Reto Freuler
   - Exponential population model with population growth rate varying each year
   - Log-normal observations of population size.
   - Parameters to estimate: mean and variance of population growth rate; observation variance.
   - Latent states: True population size in each year.
   - The example also generates predictions through 2015.

### BUGS \& NIMBLE code for the model

```{r model-code}
library(nimble)
ssmCode <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma2.proc <- pow(sigma.proc, 2)
tau.proc <- pow(sigma.proc, -2)
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process
sigma2.obs <- pow(sigma.obs, 2)
tau.obs <- pow(sigma.obs, -2)

# Likelihood
# State process
for (t in 1:(T-1)){
   r[t] ~ dnorm(mean.r, tau.proc)
   logN.est[t+1] <- logN.est[t] + r[t]
   }
# Observation process
for (t in 1:T) {
   y[t] ~ dnorm(logN.est[t], tau.obs)
   }

# Population sizes on real scale
for (t in 1:T) {
   N.est[t] <- exp(logN.est[t])
   }
})
```

Argument to `nimbleCode` was copied and pasted from the book's code.

[^1]: 
Marc K&eacute;ry and Michael Schaub. 2012. Bayesian Population Analysis Using WinBUGS: A hierarchical perspective.  Elsevier / Academic Press.

# Reading BUGS code from a file

A typical idiom for users of the R2WinBUGS and rjags packages is to use `sink` and `cat` to write BUGS code to a file from R:
```{r, ssm-cat}
sink("ssm.bug")
cat("
model {
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma2.proc <- pow(sigma.proc, 2)
tau.proc <- pow(sigma.proc, -2)
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process
sigma2.obs <- pow(sigma.obs, 2)
tau.obs <- pow(sigma.obs, -2)

# Likelihood
# State process
for (t in 1:(T-1)){
   r[t] ~ dnorm(mean.r, tau.proc)
   logN.est[t+1] <- logN.est[t] + r[t]
   }
# Observation process
for (t in 1:T) {
   y[t] ~ dnorm(logN.est[t], tau.obs)
   }

# Population sizes on real scale
for (t in 1:T) {
   N.est[t] <- exp(logN.est[t])
   }
}
",fill = TRUE)
sink()
```

This can be read into NIMBLE using `readBUGSmodel`:

```{r read-bugs-model}
ssmCode <- readBUGSmodel('ssm.bug', dir ="", returnModelComponentsOnly = TRUE)$model
file.remove('ssm.bug')
```

