---
title: "Building a model (state-space model example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```
```{r loadnimble, include=FALSE}
library(nimble)
```

# A basic example: population state-space model

We'll use an example from Chapter 5 of K&eacute;ry and Schaub
(2012)\footnote{Marc K&eacute;ry  and Michael Schaub. 2012. Bayesian Population Analysis Using WinBUGS: A hierarchical perspective.  Elsevier / Academic Press.}

Code and data are available from the [BPA web site](http://www.vogelwarte.ch/de/projekte/publikationen/bpa/complete-code-and-data-files-of-the-book.html), in bpa-code.txt.

### Summary

   - House martin counts (1990 - 2009) by Reto Freuler
   - Exponential population model with population growth rate varying each year
   - Log-normal observations of population size.
   - Parameters to estimate: mean and variance of population growth rate; observation variance.
   - Latent states: True population size in each year.
   - The example also generates predictions through 2015.

### BUGS \& NIMBLE code for the model

This has been simplified (without changing the model) from the book's code files.

```{r model-code}
library(nimble)
ssmCode <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process

# Likelihood
# State process
for (t in 1:(T-1)){
   r[t] ~ dnorm(mean.r, sd = sigma.proc)
   logN.est[t+1] <- logN.est[t] + r[t]
   }
# Observation process
for (t in 1:T) {
   y[t] ~ dnorm(logN.est[t], sd = sigma.obs)
   }

# Population sizes on real scale
for (t in 1:T) {
   N.est[t] <- exp(logN.est[t])
   }
})
```

# Building a model in NIMBLE

   - WinBUGS/OpenBUGS/JAGS: provide code, data and (optional) initial values at one time, and the software runs an MCMC.
   - NIMBLE gives more control: data is different from constants and can be set and modified later.  Model is separate from algorithms.

Building the state-space model in one step:
```{r build-model}
# Code from BPA book:
pyears <- 6 # Number of future years with predictions
hm <- c(271, 261, 309, 318, 231, 216, 208, 226, 195, 226, 233, 209, 226, 192, 191, 225,
        245, 205, 191, 174, rep(NA, pyears))
year <- 1990:(2009 + pyears)

# Bundle data
bugs.data <- list(y = log(hm), T = length(year))
## NIMBLE will handle y as data, T as a constant

# Initial values
inits <- function(){list(sigma.proc = runif(1, 0, 1), mean.r = rnorm(1),
                         sigma.obs = runif(1, 0, 1), logN.est = c(rnorm(1, 5.6, 0.1),
                                                         rep(NA, (length(year)-1))))}

ssm <- nimbleModel(ssmCode, constants = bugs.data, inits = inits()) # inits handled differently 
```

### `nimbleModel` returns an object with which you can program.
   - This will be covered later.

# Compiling a model

In general, you'll want a version of the model that allows for fast computation (this can then be used by any algorithms you use on the model).

To create a fast compiled version of the model, you simply do this.

```{r, compile-model}
Cssm <- compileNimble(ssm)
```

# Looking at the model's graph

A central concept for hierarchical statistical modeling is:

   - Think of the model as a graph, typically a directed acyclic graph
   (DAG).
   - Every line of BUGS code declares a node on the graph.

We can harness NIMBLE's use of the igraph library to plot the graph.

```{r, plot-graph, fig.cap=""}
## Remove N.est[t] to reduce clutter 
ssmCode <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process

# Likelihood
# State process
for (t in 1:(T-1)){
   r[t] ~ dnorm(mean.r, sd = sigma.proc)
   logN.est[t+1] <- logN.est[t] + r[t]
   }
# Observation process
for (t in 1:T) y[t] ~ dnorm(logN.est[t], sd = sigma.obs)

})

# Make a smaller version so the graph wil be readable
ssmSmall <- nimbleModel(ssmCode, constants = list(T = 5),
                        data = list(y = bugs.data$y[1:5]))
library(igraph)
graph <- ssmSmall$getGraph()
plot(graph, layout = layout_(graph, with_kk())) ## uses plot.igraph
```

igraph's plot function doesn't know how to lay out a state-space model
but this looks ok.
