---
title: "Customizing and comparing MCMCs (state-space example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```
```{r loadnimble, include=FALSE}
library(nimble)
read_chunk('chunks_ssm.R')
```

# Outline

In this module we will:

   - Use the house martin state-space model from K&eacute;ry and Schaub (2012).
   - Try block samplers for the `r[t]`s.  Often a key factor that
     reduces MCMC performance is dependence between parameters that
     limits the ability of univariate samplers to move very far. A
     standard strategy is to sample correlated parameters in blocks.

       + This actually doesn't help much in this model (and will
       be more efficient in a future NIMBLE release) but we'll
       illustrate it anyway.
	   
   - Try Metropolis-Hastings samplers on a log scale for `sigma.obs` and `sigma.proc`.
   - Automatically compare a customized MCMC to the default MCMC using `compareMCMCs`.

# Build the model and default MCMC

We first need to build the model.

```{r, ssm-code}
```

```{r, ssm-model}
```

```{r, ssm-compile}
```

```{r, ssm-mcmc}
ssmMCMC <- buildMCMC(ssm) # can skip mcmcConf
CssmMCMC <- compileNimble(ssmMCMC, project = ssm)
CssmMCMC$run(10000)
samples <- as.matrix(CssmMCMC$mvSamples)
tsplot <- function(x, ...) plot(seq_along(x), x, type = 'l', ...)

par(mfrow = c(1, 3), mai = c(.6, .5, .1, .2))
tsplot(samples[ , 'mean.r'], xlab = 'iteration',
     ylab = 'mean.r', main = 'mean.r')
tsplot(samples[ , 'sigma.proc'], xlab = 'iteration',
     ylab = 'sigma.proc', main = 'sigma.proc')
tsplot(samples[ , 'sigma.obs'], xlab = 'iteration',
     ylab = 'sigma.obs', main = 'sigma.obs')
```

# Customizing and MCMC configuration

MCMC is a family of algorithms and there are many ways to create an MCMC for any given model by choosing different samplers (or "updaters").  Different choices can have vastly different efficiencies for different models and data sets.

There are several ways we can customize an MCMC configuration in NIMBLE.

Here we will start from a default configuration and then remove and add samplers.

```{r, ssm-mcmc2}
mcmcConf2 <- configureMCMC(ssm) 
mcmcConf2$removeSamplers('sigma.obs')
mcmcConf2$removeSamplers('sigma.proc')
mcmcConf2$removeSamplers('r')
mcmcConf2$addSampler(target = 'sigma.obs', type = 'RW', control = list(log=TRUE))
mcmcConf2$addSampler(target = 'sigma.proc', type = 'RW', control = list(log=TRUE))
## illustrate how we can access all r[t] nodes:
ssm$expandNodeNames('r')
mcmcConf2$addSampler(target = ssm$expandNodeNames('r'), type = 'RW_block')
mcmcConf2$printSamplers()
```

# Build, compile and run the customized MCMC

```{r, run-ssm-mcmc2}
ssmMCMC2 <- buildMCMC(mcmcConf2)
Cssm <- compileNimble(ssm)
CssmMCMC2 <- compileNimble(ssmMCMC2, project = ssm)
CssmMCMC2$run(10000)
```

```{r}
samples2 <- as.matrix(CssmMCMC2$mvSamples)
tsplot <- function(x, ...) plot(seq_along(x), x, type = 'l', ...)

par(mfrow = c(1, 3), mai = c(.6, .5, .1, .2))
tsplot(samples[ , 'mean.r'], xlab = 'iteration',
     ylab = 'mean.r', main = 'mean.r')
tsplot(samples[ , 'sigma.proc'], xlab = 'iteration',
     ylab = 'sigma.proc', main = 'sigma.proc')
tsplot(samples[ , 'sigma.obs'], xlab = 'iteration',
     ylab = 'sigma.obs', main = 'sigma.obs')

```

# Comparing MCMC configurations

We've made two kinds of MCMCs for the same model.  How can we compare
them?

We must consider two factors:

   - How well the MCMC mixes.  This is measured by the *effective
      sample size* (ESS) for each parameter.

   - How quicly the MCMC runs.  This is measured by its computation
   time.

We define *MCMC efficiency* for one parameter as

   - MCMC efficiency (of a parameter) = ESS / time (in seconds)

That gives a measure of performance for each parameter, but how we do
summarize those across parameters to get a single number?

Often the trustworthiness of results is limited by the worst-mixing
parameter.  Therefore we define

   - MCMC efficiency (of an algorithm) = minimum MCMC efficiency of
     all 

# Comparing MCMC configurations using `compareMCMCs`

After some preliminary exploration, one sees that the slowest-mixing
parameters are the two standard deviations, and a good strategy for
them to mix is to include two samplers for each one on a regular scale
and one on a log scale. So we'll include just that modification for
the comparison.

```{r, compare-mcmcs}
set.seed(1)
modelInfo <- list(ssm = list(code = ssmCode, data = bugs.data['y'], constants = bugs.data['T'], inits = inits()))
MCMCdefs <- list(custom = quote({
    mcmcConf2 <- configureMCMC(Rmodel) # can skip mcmcConf
    mcmcConf2$addSampler(target = 'sigma.obs', type = 'RW', control = list(log=TRUE))
    mcmcConf2$addSampler(target = 'sigma.proc', type = 'RW', control = list(log=TRUE))
    mcmcConf2
}))
comparisonResults <- compareMCMCs(modelInfo = modelInfo,
                                  MCMCs = c('nimble', 'custom'),
                                  MCMCdefs = MCMCdefs,
                                  niter = 50000,
                                  burnin = 5000, summary = FALSE)
```

Make comparison figures

```{r, make-comparison-figures}
make_MCMC_comparison_pages(comparisonResults, dir = 'ssm_custom_MCMC_results')
## open ssm_custom_MCMC_results/ssm.html in a browser to see results
## The name file name ssm is taken from the name in the modelInfo list
```

The results are [here](ssm_custom_MCMC_results/ssm.html)
