---
title: "How NIMBLE extends BUGS and JAGS (and what is missing)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

NIMBLE extends the BUGS language in many ways, and it still has some important limitations (which we hope to overcome).

# How NIMBLE extends BUGS

  - Alternative distribution parameterizations (like R).
  - Named parameters (like R).
  - Vectorized math and linear algebra
  - Definition-time if-then-else (multiple model variants from the same code)
  - User-defined functions and distributions

# How NIMBLE is limited compared to BUGS and/or JAGS:

   - NIMBLE doesn't handle stochastic indices, such as in models with latent group membership.
   - NIMBLE is stricter about requiring square brackets and informative indices for non-scalar nodes.


# Alternative distribution parameterizations and named parameters

In BUGS and JAGS, there is only one parameterization for each distribution.  
```{r, eval=FALSE}
x ~ dnorm(0, .01)
```
means `x` has a mean of 0 and a **precision** of .01. (Precision = 1/variance).

In NIMBLE, the *default* parameterization follows BUGS and JAGS, but there are alternatives:
```{r, eval = FALSE}
x ~ dnorm(0, sd = 10)
```
means 'x' has a mean of 0 and a **standard deviation** of 10.

#