---
title: "How NIMBLE extends BUGS and JAGS (and what is missing)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```
```{r loadnimble, include=FALSE}
library(nimble)
```

NIMBLE extends the BUGS language in many ways, and it still has some important limitations (which we hope to overcome).

# How NIMBLE extends BUGS

  - Alternative distribution parameterizations (like R).
  - Named parameters (like R).
  - Vectorized math and linear algebra
  - Definition-time if-then-else (multiple model variants from the same code)
  - User-defined functions and distributions

# How NIMBLE is limited compared to BUGS and/or JAGS:

   - NIMBLE doesn't handle stochastic indices, such as in models with latent group membership.
   - NIMBLE is stricter about requiring square brackets and informative indices for non-scalar nodes.


# Alternative distribution parameterizations and named parameters

In BUGS and JAGS, there is only one parameterization for each distribution.  
```{r, eval=FALSE}
x ~ dnorm(0, .01)
```
means `x` has a mean of 0 and a **precision** of .01. (Precision = 1/variance).

In NIMBLE, the *default* parameterization follows BUGS and JAGS, but there are alternatives:
```{r, eval = FALSE}
x ~ dnorm(0, sd = 10)
```
means 'x' has a mean of 0 and a **standard deviation** of 10.

# Vectorized math and linear algebra

In BUGS and JAGS, if there are a parallel set of scalar declarations, each must be made separately:
```{r, eval = FALSE}
for(i in 1:10) {
  logit(p[i]) <- intercept + slope * x[i]
}
```

In NIMBLE this can be made as a vector declaration:

```{r, eval = FALSE}
   logit(p[1:10]) <- intercept + slope * x[1:10]
```

*Whether this is advantageous depends on the algorithm that will use the model.*  In the second version, `p[1:10]` will *always* be calculated as a vector.

Vectorization *can be good* if `x[1:10]` is constant, because any new value of `intercept` or `slope` will require calculation of all values of `p[i]` anyway, so it may be more efficient to calculate them all together.

Vectorization *can be bad* if each `x[i]` may change, such as if they are random effects. If `x[i]` changes, only the corresponding `p[i]` should need to be calculated.

Determining the best approach requires *thinking about the graph.*  In the first case, there are 10 scalar nodes, named "`p[1]`" - "`p[10]`".  In the second case, there is one vector node, named "`p[1:10]`".

# Definition-time if-then-else

In BUGS and JAGS, each model variant requires its own code.  This leads to lots of copying and pasting of code, the opposite of modular programming with code re-use.

In NIMBLE, if-then-else statements will be evaluated *in R, when the model is defined* to determine what code to include in the model.

Example:
```{r, eval=FALSE}
regressionCode <- nimbleCode({
    intercept ~ dnorm(0, sd = 1000)
    slope1 ~ dnorm(0, sd = 1000)
    if(includeX2) {  ## includeX2 could be any R expressions
    ## This clause has code for including x2
        slope2 ~ dnorm(0, sd = 1000)
        for(i in 1:N) 
            predictedY[i] <- intercept + slope1 * x1[i] + slope2 * x2[i]
    } else {
    ## This clause has code for omitting x2
        for(i in 1:N) predictedY[i] <- intercept + slope1 * x1[i]
    }
    sigmaY ~ dunif(0, 100)
    for(i in 1:N) Y[i] ~ dnorm(predictedY[i], sigmaY)
})

includeX2 <- FALSE ## includeX2 is defined in the R global environment
modelWithoutX2 <- nimbleModel(regressionCode, constants = list(N = 30), 
                              check=FALSE)

includeX2 <- TRUE
modelWithX2 <- nimbleModel(regressionCode, constants = list(N = 30), 
                           check = FALSE)
```

# User-defined functions and distributions

These will be introduced in a separate module.
