---
title: "Programming with models (2): Inspecting model structure (state-space model example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r loadnimble, include=FALSE}
library(nimble)
```
```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
read_chunk('chunks_ssm.R')
```
# Outline

In this module, we'll learn how to query a model about its graph.

*model-generic programming = querying a model's graph + operating a model*

Specifically, we'll learn:

   - about topological ordering of nodes and why it matters.
   - how to determine the dependencies (or children) of one or more nodes, and
       + what different kinds of dependencies mean.
   - how to expand node names.
   - how to determine node categories such as *top-level nodes, latent-state nodes, and data nodes*.
   - how NIMBLE sometimes inserts *lifted nodes* into a model and what that means for you.
   - how to determine the variables in a model.

We first need to build the model.  For this module we will use the smaller version of the state-space model, with only the first four time points.  The only purpose is to simplify the examples and the graph figures for clarity.

```{r, ssmSmall-code}
```

```{r, ssmSmall-model}
```

# Topological ordering of nodes and why it matters.


```{r}
library(nimble)
library(igraph)
ssmSmallCode <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process

# Likelihood
# State process
for (t in 1:(T-1)){
   r[t] ~ dnorm(mean.r, sd = sigma.proc)
   logN.est[t+1] <- logN.est[t] + r[t]
   }
# Observation process
for (t in 1:T) y[t] ~ dnorm(logN.est[t], sd = sigma.obs)

})

## @knitr ssmSmall-model
pyears <- 6 # Number of future years with predictions
hm <- c(271, 261, 309, 318, 231, 216, 208, 226, 195, 226, 233, 209, 226, 192, 191, 225,
        245, 205, 191, 174, rep(NA, pyears))
year <- 1990:(2009 + pyears)

# Bundle data
bugs.data <- list(y = log(hm), T = length(year))
## NIMBLE will handle y as data, T as a constant

ssmSmall <- nimbleModel(ssmSmallCode, constants = list(T = 5),
                        data = list(y = bugs.data$y[1:5]))

drawGraph <- function(model, colorBy = "none", colors = c('salmon','cyan','plum')) {
    graph <- model$getGraph()
    numNodes <- length(ssmSmall$getNodeNames())
    vertex.color <- rep('grey', numNodes)
    if(identical(colorBy, "none")) {}
    else if(length(colorBy) > 1) {
        if(is.character(colorBy))
            colorBy <- model$expandNodeNames(colorBy, returnType = "ids")
        vertex.color[colorBy] <- colors[1]
    } else if(colorBy == "stochDetermData") {
        stochIDs <- ssmSmall$getNodeNames(stochOnly = TRUE, returnType = "ids")
        determIDs <- ssmSmall$getNodeNames(determOnly = TRUE, returnType = "ids")
        dataIDs <- ssmSmall$getNodeNames(dataOnly = TRUE, returnType = "ids")
        vertex.color[stochIDs] <- colors[1]
        vertex.color[dataIDs] <- colors[2]
        vertex.color[determIDs] <- colors[3]
    } else if(colorBy == "topLatentEnd") {
        topIDs <- ssmSmall$getNodeNames(topOnly=TRUE, returnType = "ids")
        latentIDs <- ssmSmall$getNodeNames(latentOnly=TRUE, returnType = "ids")
        endIDs <- ssmSmall$getNodeNames(endOnly=TRUE, returnType = "ids")
        vertex.color[topIDs] <- colors[1]
        vertex.color[latentIDs] <- colors[2]
        vertex.color[endIDs] <- colors[3]
    }
    plot(graph, layout = layout_(graph, with_kk()),
         vertex.color = vertex.color ) ## uses plot.igraph
}
```

Suppose we want to change the value of `r[3]` and update all relevant
parts of the graph.  The nodes that need to be updated are shown in red.


```{r}
drawGraph(ssmSmall, ssmSmall$getDependencies('r[3]'))
```

Note that **order matters**.  If we calculated the log probability
density of `y[4]` *before* calculating `logN.est[4]`, then `y[4]` would
use the *old* value of `logN.est[4]`.  That would be a mistake.

A valid order for calculations is called *topologically sorted*.  Most
of NIMBLE's model-querying functions will return nodes in tologically sorted order.

# Inspecting the model: Querying graph dependencies

There are several kinds of dependencies that may be needed by different algorithms.

   - *Stochastic dependencies* (called *Markov blanket* in computer science) terminate at stochastic nodes (inclusive)

```{r}
ssmSmall$getDependencies('r[3]')
```
```{r, echo=FALSE}
drawGraph(ssmSmall, ssmSmall$getDependencies('r[3]'))
```
       + These are the dependencies needed for a Metropolis-Hastings
       MCMC sampler.

   - *Deterministic dependencies* terminate at stochastic nodes but
omit the stochastic nodes
```{r}
ssmSmall$getDependencies('r[3]', determOnly = TRUE)
```
```{r, echo=FALSE}
drawGraph(ssmSmall, ssmSmall$getDependencies('r[3]', determOnly = TRUE))
```

   - Other kinds of control can be found by
```{r}
args(ssmSmall$getDependencies)
## or help(modelBaseClass)
```


# Finding out about nodes

The functions `model$getNodeNames`, `model$expandNodeNames`,
`model$topologicallySortNodes` will query the model about nodes and
their roles in the model.

Examples include:

   - All nodes in topologically-sorted order
```{r}
ssmSmall$getNodeNames()
```

   - Stochastic, deterministic, and data nodes (which are also
   stochastic).

```{r}
ssmSmall$getNodeNames(stochOnly = TRUE)   #salmon
ssmSmall$getNodeNames(determOnly = TRUE)  #cyan
ssmSmall$getNodeNames(dataOnly = TRUE)    #plum
```
```{r}
drawGraph(ssmSmall, 'stochDetermData')
```

   - Top nodes, latent nodes, and end nodes

```{r}
ssmSmall$getNodeNames(topOnly = TRUE)   #salmon
ssmSmall$getNodeNames(latentOnly = TRUE)  #cyan
ssmSmall$getNodeNames(endOnly = TRUE)    #plum
```
```{r}
drawGraph(ssmSmall, 'topLatentEnd')
```

Another kind of node is called *RHSonly* if it only appears on the
right-hand side of a BUGS declaration and never on the left-hand side.

You can get the nodes present in some arbitrary range of a
variable like this:
```{r}
ssmSmall$expandNodeNames('logN.est[2:5]')
```

Finally, if for some reason you did not obtain a vector of node names
from the model in the first place and need it topologically sorted,
you can do that:

```{r}
ssmSmall$topologicallySortNodes(c('logN.est[5]', 'r[4]', 'logN.est[3]', 'y[2]'))
```

Often there is more than one valid topological sorting, but NIMBLE
chooses one and uses it for your model.

# The importance of querying a model: lifted nodes.

*You should never assume you know what nodes are in a model simply
because you wrote the BUGS code.*

NIMBLE implements some features by inserting its own nodes.

Let's look at the two main ways this happens:

```{r}
m1 <- nimbleModel(
    nimbleCode({
        tau ~ dunif(0, 100)
        x ~ dnorm(0, tau) #by default, tau is a precision
    }))
plot(m1$getGraph())
m1$getNodeNames()
```
The node `lifted_d1_over_sqrt_oPtau_cP` has been inserted between
`tau` and `x`.   The resulting model would equivalently have been
created by this BUGS code:

```{r, eval = FALSE}
nimbleCode({
    tau ~ dunif(0, 100)
    lifted_d1_over_sqrt_oPtau_cP <- 1/sqrt(tau)
    x ~ dnorm(0, sd = lifted_d1_over_sqrt_oPtau_cP) #by default, tau is a precision
}))
```

NIMBLE has *lifted* the calculation of standard deviation from
precision so that it is part of the model's graph.  Therefore *you
will make a mistake if you assume that the dependencies of `tau`
include only `x`*:

```{r, eval = FALSE}
m1$calculate(c('tau','x'))
```

The correct way to do this is with model-generic programming:
```{r, eval = FALSE}
m1$calculate( m1$getDependencies('tau') )
```

The second common situation that creates lifted this is:
```{r}
m2 <- nimbleModel(
    nimbleCode({
        a ~ dnorm(0, 1)
        b ~ dnorm(a + 1, 1)
    }))
plot(m2$getGraph())
m2$getNodeNames()
```

In this case, NIMBLE lifted the expression `a+1` and created a node
for it.   Again the correct way to program is to query a model to
learn about its nodes.  Do not assume you know what the nodes are.

# Querying the sizes and types of model variables

Sometimes it is useful to ask a model about its *variables*  and their
sizes.  This information is not competely nicely arranged, but it is
available:

```{r}
ssmSmall$getVarNames()
ssmSmall$getVarInfo('r')
```
