---
title: "Programming with models (2): Inspecting model structure"
subtitle: "NIMBLE training materials module"
author: "Nimble Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)  # otherwise new() not being found - weird
library(nimble)
read_chunk("chunks.R")
```

# Running example: the pump model

We first need to build the model.

```{r, pump-code}
```

```{r, pump-model}
```

```{r, pump-compile}
```

# Nodes and variables

The parameters and data in a model are represented as nodes in a graph. Here we define some terms:

 - parameter: an unknown quantity in the model that is represented as a random variable and will generally be estimated in an algorithm, e.g. $\alpha$ in the pump model
 - data: a known quantity in a model, also represented as a random variable, e.g., $x_1, x_2, \ldots$ in the pump model
 - constants: other fixed quantities involved in the model, e.g., $N$ in the pump model
 - node: an element in the model representing data, parameter, or a deterministic quantity that is a function of other quantities in the model, e.g., $\alpha$ or $x_3$ in the pump model
 - variable: a collection of one or more nodes with the same name, e.g., $\alpha$ or $x$ in the pump model 

# Inspecting the model

Nimble provides users and programmers with the ability to get information about the nodes and variables in the model and the relationships amongst them.

What are the variables in the model? What are the dependencies of a given node (this is information used in many algorithms)?

```{r, model-nodes}
pump$getNodeNames()
hypers <- pump$getNodeNames(topOnly = TRUE)
hypers
pump$getDependencies(hypers)
```

