---
title: "Programming with models (3): Writing nimbleFunctions"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)  # otherwise new() not being found - weird
library(nimble)
```

# Introduction

*nimbleFunctions* are at the heart of NIMBLE. They are the way that algorithms are implemented. They can also be used for user-defined functions and samplers for use in BUGS code (see the [user_function](user_function.html) and [user_sampler](user_sampler.html) modules) and they can be used to compile parts of R (see the *compile_R* module), without reference to a model. But their main purpose is providing a way for developers to implement algorithms.

# Components of a nimbleFunction

NIMBLE uses the concept of *two-stage evaluation* from computer science to run a model-specific algorithm based on model-generic algorithm code. The first stage of evaluation specializes the algorithm to the model of interest via *setup* code. The second stage runs the algorithm via *run* code. 

Thus, a nimbleFunction has two parts:

 - setup code: used to tailor the algorithm to a particular model structure. Often this involves determining dependencies amongst nodes in the model and setting up storage using modelValues
 - run code: the guts of the algorithm, written generically so it will apply to any (appropriate) model

Setup code is written as a R function, using R code, usually including NIMBLE's special functions for querying the model structure (see the module *model_structure*). 

Run code is written using the NIMBLE DSL. While this is formally a language distinct from R, you can just think of it as a subset of R, enhanced with some functions for operating on the model (see the module *operating_model*). 

# Some syntax for nimbleFunctions

Here are some of the functions you may use in the *run* function of a nimbleFunction:

 - *returnType*, e.g. ```returnType(double(1))``` for a vector of reals;
 - *length*, e.g., ```length(x)``` to determine the length of a run-time argument *x*;
 - *calculate*, *simulate*, *getLogProb* to manipulate the model;
 - *values* and *nimCopy* to copy values;
 - *print*;
 - functions to declare multivariate variables such as *numeric*, *matrix*, and *array*; and
 - random number generation functions, e.g., ```rnorm(1, 100, 5)``` (these must always only generate a single random variable).

Section 9 of the NIMBLE User Manual describes the syntax for *run* code in detail, including lots of neat functionality such as using nested nimbleFunctions and having multiple run-time functions (i.e., methods) as part of a nimbleFunction.


# A basic example

Let's consider how we would optimize the parameters in a model using a nimbleFunction. Basically, we'll just construct an objective function that we can then pass to R's *optim* function to do the actual numerical optimization.

This amounts to setting things up to find the posterior mode of a model; this is generally a reasonable thing to do only for models with a small number of parameters and without hierarchical structure.

Here's the generic nimbleFunction that could be used with any model.

```{r, nf}
objective <- nimbleFunction(
    setup = function(model, target) {
          calcNodes <- model$getDependencies(target)
          },
    run = function(par = double(1)) {
        returnType(double(0))
        values(model, target) <<- par
        ans <- model$calculate(calcNodes)
        return(ans)
    }
)
```

# Using the nimbleFunction on a model

Let's try to optimize the parameters of GLM, in particular a basic logistic regression, for which the maximum likelihood estimates are not available in closed form.

We'll build the model and then create a nimbleFunction specific to the actual model. And we'll create a compiled version of that nimbleFunction to use in the actual optimization.

```{r, model}
code <- nimbleCode({
     for(i in 1:n) {
           y[i] ~ dbern(p[i])
           p[i] <- expit(b0 + b1*x[i])
     }
})
n <- 300
b0 <- -0.5
b1 <- 4
x <- runif(n, -1, 1)
p <- expit(b0 + b1*x)
y <- rbinom(n, 1, p)

model <- nimbleModel(code, constants = list(n = n, x = x),
      data = list(y = y), inits = list(b0 = 0, b1 =0))
Cmodel <- compileNimble(model)
logistic_obj <- objective(model, c('b0', 'b1'))
Clogistic_obj <- compileNimble(logistic_obj, project = model)
```

Now let's run the optimization and compare to R's *glm* function. 

```{r, optimizer}
out <- optim(c(Cmodel$b0, Cmodel$b1), Clogistic_obj$run, 
    control = list(fnscale = -1))
out$par
glm(y ~ x, family = 'binomial')
```

Looks like it works.

# Example of a simple MCMC sampler      

Here's a simplified version of the random walk sampler used for NIMBLE's MCMC engine. Note that this simplification excludes the use of an adaptive algorithm that adjusts the proposal standard deviation to the acceptance rate as the MCMC iterations proceed.

```{r, basic-sampler, eval=FALSE}
sampler_RW <- nimbleFunction(
    contains = sampler_BASE,  # some object-oriented inheritance for MCMC samplers
    setup = function(model, mvSaved, target, control) {
        targetAsScalar <- model$expandNodeNames(target, returnScalarComponents = TRUE)
        if(length(targetAsScalar) > 1)   
           stop('cannot use RW sampler on more than one target; try RW_block sampler')
        if(model$isDiscrete(target))     
           stop('cannot use RW sampler on discrete-valued target; try slice sampler')
        ###  control list extraction  ###
        scale <- control$scale
        ###  node list generation  ###
        calcNodes  <- model$getDependencies(target)
    },

    run = function() {
        currentValue <- model[[target]]
        # note that scale is available because created in setup code
        propValue <- rnorm(1, mean = currentValue,  sd = scale)
        model[[target]] <<- propValue
        logMHR <- calculateDiff(model, calcNodes) 
        jump <- decide(logMHR)
        if(jump) nimCopy(from = model, to = mvSaved, row = 1, 
                         nodes = calcNodes, logProb = TRUE)
        else     nimCopy(from = mvSaved, to = model, row = 1, 
                         nodes = calcNodes, logProb = TRUE)
    },

    where = getLoadingNamespace()
)
```


