---
title: "Programming with models (1): Operating a model (state-space model example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r loadnimble, include=FALSE}
library(nimble)
```
```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
read_chunk('chunks_ssm.R')
```

# Outline

In this module we'll see how a model can be operated.  This will
include:

   - The distinction between nodes and variables.
   - NIMBLE's node naming convention
   - Getting and setting values.
   - Ability to do two core operations:
      + *calculate()* 
      + *simulate()* 
   - Ability to do three other helpful operations
      + *getLobProb()* 
      + *calculateDiff()*
      +  *getParam()* 
   - More general getting and setting of values
      + *values()*
   - Equivalent operation of uncompiled and compiled models

#Running example: state-space model example

We first need to build the model.

```{r, ssm-code}
```

```{r, ssm-model}
```

```{r, ssm-compile}
```

# Nodes and variables

The parameters, latent states, and data in a model are represented as nodes in a graph.  We'll cover more about the graph later.

An important distinction is between *nodes* and *variables*.

 - node: an element in the model that will always be calculated as a unit, such as "`r[2]`" or "`y[4]`":
    + nodes represent a graph "vertex" (or simply "node").
    + a deterministic node is calculated by updating its value
    + a stochastic node is calculated by updating the log probability density of its value.
 - variable: a collection of one or more nodes with the same name, e.g., `r` or `y` in the state-space model 

# NIMBLE's node naming convention

Every node has a character name.  In our example these are:
```{r node-names}
ssm$getNodeNames()
```

In a model with scalar nodes embedded in multivariate variables, we
would have names like `x[2, 3]` or `z[3, 2, 5]`.

In a model with multivariate nodes we would have names like
`a[1:5, 2]` or `b[2:5, 3:6]`.

*The spaces after the commas are important*, but you will usually 
 obtain node names from a model object, so they will be correct. (You can use `model$expandNodeNames` to ensure they are correctly formatted.)

We will see more about querying the model about its nodes and
variables later.

# Values in the model

What's the current value in a node or the value of the log probability density for that node?

```{r, model-values}
ssm$mean.r
ssm$logProb_mean.r  # log-likelihood components can be seen like this, but 
                    # usually should be obtained by ssm$getLogProb('mean.r'); see below. 
ssm$r
ssm$r <- rnorm(length(ssm$r), 0, 1)
ssm$r
ssm$logProb_r # not calculated yet!
```

# Operating a model: *calculate()* 

When you program with a model, you are in charge of using the model
correctly.  NIMBLE's programming system provides tools to help you do
so, but it does not try to read your mind.

*Calculate* means:

   - For a stochastic node, calculating the log probability (density)
    and storing it as the "logProb" value.
   - For a deterministic node, calculating the node's value

Calling `model$calculate(nodes)` will

   - Calculate each element of `nodes` **in order**.
   - Return the sum of the log probabilities
   - The log probability contribution from a deterministic node is 0.

Let's see an example:

```{r, model-operate}
ssm$r[1:3] <- c(.1, .2, .3)
ssm$calculate(c('r[1]','r[2]','r[3]')) 
ssm$logProb_r[1:3]
sum(ssm$logProb_r[1:3])
## The following is equivalent to what we just did:
## ssm$calculate('r[1:3]')
## If we wanted to calculate all of r, we could do:
## ssm$calculate('r')
```

Note that values of `logN.est` are *out of date* with the new values
of `r[1:3]`.  We would need to call `calculate` for `logN.est` to
update them.  The model will not read your mind about whether you want
to update those values or not; you are responsible for using the model
correctly.  As we will see later, you can query the model itself to make sure you do so.

# Operating a model: *simulate()*
 
 *Simulate* means
 
   - for a stochastic node: draw a random number and store it as
    the node's value
   - for a deterministic node: calculate and store the value (the
    same as calculate).

Calling `simulate$model(nodes)` will:

   - Simulate each element of `nodes` **in order**. 
   - return nothing.

If you want to see the new values, you must look at the model's variables.

```{r, model-simulate}
set.seed(0)  # so the calculations are reproducible
ssm$simulate('r[1:3]')  # simulate from prior
ssm$r[1:3]
```
You must call `ssm$calculate('r[1:3]')` if you want the log probability values updated to correspond to the new values of `r[1:3]`.  Again, you are in control of the model.

# Operating a model: getParam()

*getParam* means:

   - obtain the value of a parameter to a distribution.

Example:

```{r}
ssm$getParam('r[1]', 'sd')
```

You can get the value of alternative parameters that you didn't even
use in the BUGS code:

```{r}
ssm$getParam('r[1]', 'tau') # tau is the name for precision
```

`getParam` can only be used for one node at a time.

# Operating a model: getLogProb and calculateDiff

`model$getLogProb(nodes)`

   - returns the sum of stored log probability (densities) for
   elements of `nodes`
   - does **not** actually calculate the log probabilities.
   - deterministic nodes in `nodes` contribute 0 to the result.

```{r}
ssm$getLogProb('r[1:3]')
```

`model$calculateDiff(nodes)` is just like `model$calculate(nodes)` except:

   - it returns the sum of the differences between the new log probabilities and the previously stored ones.

This is often useful in Monte Carlo algorithms such as Metropolis-Hastings samplers.

# Getting and setting values of groups of variables

Sometimes it is useful to set or get model values from a single
vector of node names.  `values` does that.

```{r}
values(ssm, c('r[1:3]','logN.est[1:3]'))
values(ssm, c('r[1:3]','logN.est[1:3]')) <- 1:6
ssm$r[1:3]
ssm$logN.est[1:3]
```

# Compiled models can be used just like uncompiled models

```{r}
set.seed(0)
Cssm$r[1:3]
Cssm$simulate('r[1:3]')
Cssm$calculate('r[1:3]')
Cssm$r <- c(1,2,3)
Cssm$calculate('r[1:3]')
```

# Putting the pieces together

Say we want to put new values in `r[1:3]` and update all values and
log probabilities.  This takes several steps:

```{r}
## Put new values in r[1:3]
ssm$r[1:3] <- c(0.1, 0.2, 0.3)
## need to update log probabilities of r[1:3], values of logN.est[2:T],
## and log probabilities of logN[2:T], in that order.
## If we go out of order, calculations will not be correct.
ssm$calculate(c('r[1:3]','logN.est[2:T]', 'y[2:T]'))
```

What we really should want to do is determine which nodes to calculate, in a valid order, from the model itself.  That will allow us to write *model-generic* code.  We'll get to that later.
