---
title: "Setting up a particle filter and Particle MCMC"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---
```{r loadnimble, include=FALSE}
library(nimble)
```
```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```

# Overview

Particle filters (or Sequential Monte Carlo) are a method for simulated approximation of the likelihood of a state-space model.

NIMBLE provides a basic ("bootstrap") PF, an Auxiliary PF, and a Liu-West filter.  The last one attempts to filter parameters as well and is not an algorithm that always works well.  Of course these are written as nimbleFunctions so you can look at the R code and then create your own new idea.

In NIMBLE one has the flexibility to compose a Particle MCMC (PMCMC).  This method uses MCMC for parameters and a particle filter to evaluate the state-space likelihood needed by the MCMC. Andrieu et al. proved this is valid.

When MCMC of parameters and states works well, it will typically be more efficient that PMCMC.  But there is a lot of current research on PMCMC for models with it is not simple to set up MCMC for latent states.

We will show two ways to set up a PMCMC:

   1.  NIMBLE provides an MCMC sampler already set up to use random-walk Metropolis-Hastings for one or more parameters and evaluate the state-space likelihood using a PF (bootstrap or auxiliary).
   2. (Optional) NIMBLE provides an MCMC sampler that allows you to provide *any* other nimbleFunction as the likelihood evaluator.  We illustrate this by showing how it can be used to construct a PMCMC at a more fine-grained level than item 1. 

# Set up state-space model example

```{r}
ssmCodeAlt <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process

# Likelihood
# State process
for (t in 1:(T-1)){
   logN.est[t+1] ~ dnorm(logN.est[t] + mean.r, sd = sigma.proc)
   }
# Observation process
for (t in 1:T) {
   y[t] ~ dnorm(logN.est[t], sd = sigma.obs)
   }

# Population sizes on real scale
for (t in 1:T) {
   N.est[t] <- exp(logN.est[t])
   }
})

pyears <- 6 # Number of future years with predictions
hm <- c(271, 261, 309, 318, 231, 216, 208, 226, 195, 226, 233, 209, 226, 192, 191, 225,
        245, 205, 191, 174, rep(NA, pyears))
year <- 1990:(2009 + pyears)

# Bundle data
bugs.data <- list(y = log(hm), T = length(year))
## NIMBLE will handle y as data, T as a constant

set.seed(1)
inits <- function(){list(sigma.proc = runif(1, 0, 1), mean.r = rnorm(1),
                         sigma.obs = runif(1, 0, 1), logN.est = c(rnorm(1, 5.6, 0.1),
                                                         rep(NA, (length(year)-1))))}
ssm <- nimbleModel(ssmCodeAlt, constants = bugs.data, inits = inits()) 
```

# Setting up auxiliary particle filter

First we show how a particle filter can be created on its own

```{r}
## buildAuxiliaryFilter is "just" a nimbleFunction
auxFilterSSM <- buildAuxiliaryFilter(ssm, nodes = 'logN.est')
compiled <- compileNimble(ssm, auxFilterSSM)
compiled$auxFilterSSM$run(10000) ## 10000 is the number of samples
```

# Using a particle filter in PMCMC: the easy way

First we show how to use the random-walk PF sampler.  We create an empty configuration and then add a sampler for each parameter.

```{r, fig.cap = ""}
pmcmcConf <- configureMCMC(ssm, nodes = NULL,     # an empty configuration
                           monitors = c('mean.r','sigma.obs','sigma.proc')) 
pmcmcConf$addSampler(target = 'mean.r', type = 'RW_PF',
                      control = list(filterType = 'bootstrap', latents = 'logN.est'))
pmcmcConf$addSampler(target = 'sigma.proc', type = 'RW_PF',
                      control = list(filterType = 'bootstrap', latents = 'logN.est'))
pmcmcConf$addSampler(target = 'sigma.obs', type = 'RW_PF',
                      control = list(filterType = 'bootstrap', latents = 'logN.est'))
pmcmc <- buildMCMC(pmcmcConf)
Cpmcmc <- compileNimble(pmcmc, project = ssm, resetFunctions = TRUE)
Cpmcmc$run(1000)
plot(as.matrix(Cpmcmc$mvSamples)[,'mean.r'],
               xlab = 'iteration', ylab = 'mean.r')
plot(as.matrix(Cpmcmc$mvSamples)[,'sigma.obs'],
               xlab = 'iteration', ylab = 'sigma.obs')
```

It is notable how the mixing is slow until the RW samplers have had some iterations to adapt their proposal scales.

# Building a PMCMC more granularly (optional)

Next we show more of the underlying system.  We need two steps:

   1. We will add `sampler_RW_llFunction` samplers to the MCMC configuration. These will use whatever arbitrary function we provide for likelihood calculations.  NIMBLE currently has univariate and multivariate (block) random-walk Metropolis-Hastings samplers that can use arbitrary likelihood functions.  
   2. The format required by `sampler_RW_llFunction` is a nimbleFunction whose `run` function takes no arguments.  Since the auxiliary particle filter takes one argument (number of Monte Carlo samples at each time step), we will wrap it in another nimbleFunction.

# A wrapper for the PF

The purpose of this function is simply to wrap an auxiliary particle filter and provide a run function that takes no arguments.

Only for curiousity, we're wrapping an APF instead of a bootstrap (basic) PF.

```{r}
wrapAPF <- nimbleFunction(
   setup = function(model, ..., m = 10000) {
     APF <- buildAuxiliaryFilter(model, ...)
   },
   run = function() {
      returnType(double())
      return(APF$run(m))
   })

## example use:
wrapAPFssm <- wrapAPF(ssm, nodes = 'logN.est', m = 10000)
## this object won't be used further and was just shown to illustrate.
```

# RW sampler for arbitrary likelihood function

Oops, while setting up this example, we realized that our `sampler_RW_llFunction` fell out of step with another change in `sampler_RW` (which it uses) in the latest version of NIMBLe, so it is broken in the current release.  No problem: we'll just define a fixed version locally here, and you can see it:

```{r}
RW_llFunction_new <- nimbleFunction(
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control) {
        ###  control list extraction  ###
        adaptive       <- control$adaptive
        adaptInterval  <- control$adaptInterval
        scale          <- control$scale
        llFunction     <- control$llFunction
        includesTarget <- control$includesTarget
        ###  node list generation  ###
        calcNodes <- model$getDependencies(target)
        ###  nested function and function list definitions  ###
        mvInternal <- modelValues(model)
        RWControl <- list(adaptive = adaptive, adaptInterval = adaptInterval, scale = scale, logScale = FALSE, reflective = FALSE)
        targetRWSamplerFunction <- sampler_RW(model, mvInternal, target, RWControl)
        my_setAndCalculateOne <- setAndCalculateOne(model, target)
        my_decideAndJump <- decideAndJump(model, mvSaved, calcNodes)
    },

    run = function() {
        modelLP0 <- llFunction$run()
        if(!includesTarget)     modelLP0 <- modelLP0 + getLogProb(model, target)
        propValue <- rnorm(1, mean = model[[target]], sd = scale)
        my_setAndCalculateOne$run(propValue)
        modelLP1 <- llFunction$run()
        if(!includesTarget)     modelLP1 <- modelLP1 + getLogProb(model, target)
        jump <- my_decideAndJump$run(modelLP1, modelLP0, 0, 0)
        if(adaptive) {
            targetRWSamplerFunction$adaptiveProcedure(jump)
            scale <<- targetRWSamplerFunction$scale
        }
    },

    methods = list(
        reset = function() {
            targetRWSamplerFunction$reset()
        }
    )
)
```

# Make the second version of a PMCMC

```{r}
pmcmcConf2 <- configureMCMC(ssm, nodes = NULL, # an empty configuration
                            monitors = c('mean.r','sigma.obs','sigma.proc')) 
pmcmcConf2$addSampler(target = 'mean.r', type = RW_llFunction_new,
                      control = list(llFunction = wrapAPFssm, includesTarget = FALSE))
pmcmcConf2$addSampler(target = 'sigma.proc', type = RW_llFunction_new,
                      control = list(llFunction = wrapAPFssm, includesTarget = FALSE))
pmcmcConf2$addSampler(target = 'sigma.obs', type = RW_llFunction_new,
                      control = list(llFunction = wrapAPFssm, includesTarget = FALSE))
```

Compile it:

```{r}
pmcmc2 <- buildMCMC(pmcmcConf2)
Cpmcmc2 <- compileNimble(pmcmc2, project = ssm, resetFunctions = TRUE)
```

```{r, fig.cap=TRUE}
Cpmcmc2$run(1000)
plot(as.matrix(Cpmcmc2$mvSamples)[,'mean.r'],
      xlab = 'iteration', ylab = 'mean.r')
plot(as.matrix(Cpmcmc2$mvSamples)[,'sigma.obs'],
      xlab = 'iteration', ylab = 'sigma.obs')
```
