---
title: "nimbleFunction programming example: Reversible jump"
subtitle: "NIMBLE ENAR webinar module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found - weird
library(nimble)
```

# Introduction

Here we'll program another sampler using nimbleFunctions, this time applying reversible jump (RJ) MCMC for variable selection in the AFT survival model example. The nice thing about RJ for variable selection is that it is a fairly straightforward implementation of RJ.

Recall that we had two covariates in the AFT model - blood urea nitrogen and hemoglobin. Let's see if hemoglobin is needed in the model using Bayesian variable selection. 

```{r, aft-setup}
library(emplik, quietly = TRUE, warn.conflicts = FALSE)
data(myeloma)

n <- nrow(myeloma)
time <-  myeloma[ , 1]    ## survival or censoring time
vstatus <- myeloma[ , 2]  ##  0 = alive (i.e., censored)
alive <- vstatus == 0
cens_time <- rep(NA, n)
cens_time[alive] <- time[alive]
cens_time[!alive] <- Inf
time[alive] <- NA
## covariates:
logBUN <- myeloma[ , 3]
HGB <- myeloma[ , 4]
logBUN <- (logBUN - mean(logBUN)) / sd(logBUN)
HGB <- (HGB - mean(HGB)) / sd(HGB)
```

# BUGS code for reversible jump version of semiparametric AFT model

Here's the BUGS code that implements the model specified earlier, modified slightly to allow for variable selection by adding an indicator that will "turn on" or "turn off" the hemoglobin coefficient.

```{r, aft-bugs}
codeAFTsel <- nimbleCode({
    for(i in 1:n) {
        x[i] ~ dweib(alpha, lambda[i])
        is_cens[i] ~ dinterval(x[i], c[i])  ## right-censoring
        lambda[i] <- exp(eta[i] + Z[i,1]*delta[1] + indic*Z[i,2]*delta[2])
        eta[i] <- etaTilde[xi[i]]  ## mix over eta; mu = exp(eta)
    }
    indic ~ dbern(0.5)             ## turns Z2 on and off
    xi[1:n] ~ dCRP(conc, size = n) ## CRP for mixture components
    conc ~ dgamma(1, 1)
    for(i in 1:nSub)
        etaTilde[i] ~ dunif(b0, B0) ## base measure H_b
    alpha ~ dunif(a0, A0)
    for(j in 1:p)
        delta[j] ~ dflat()
})
```

# Overall strategy for variable selection via RJ

We won't go into all the details, but by looking at the rules for how to set up a valid RJ sampler, one can see that in the simple variable selection case, they simplify to straightforward calculations.

To go from the reduced model, $M^r$, to full model, $M^f$, we need to propose an auxiliary variable, $u$, that augments the dimension of the model and a transformation to produce the new coefficient, $\delta_{2}^*$, given the auxiliary variable. The simplest thing to do is:

$$ u \sim J(u) = N(0, \sigma^2) $$

$$ \delta_{2}^* = u $$

The acceptance ratio is then

$$ q = \frac{p(y|\theta^f,M^f)p(\theta^f|M^f) }{p(y|\theta^r,M^r)p(\theta^r|M^r) J(u) } $$

For those of you familiar with RJ, the lack of Jacobian here occurs because the transformation $\delta_{2}^*=u$ has Jacobian equal to 1.

To go from the full to reduced model, we have

$$ u^* = \delta_2 $$ 

and acceptance ratio:

$$ q = \frac{p(y|\theta^r,M^r)p(\theta^r|M^r)J(u^*) }{p(y|\theta^f,M^f)p(\theta^f|M^f)  } $$

# RJ sampler setup code

The setup code needs to determine the nodes involved in the full model and the reduced model, where the reduced model omits the variable under consideration.


```{r, setup-rj, eval=FALSE}
 setup = function( model, mvSaved, target, control ) {
   # target = 'indic'
   coefNode <- control$coef   # 'delta[2]'
   scale    <- control$scale  # could be made adaptive
   # with variable
   calcNodes_full <- model$getDependencies(c(coefNode, target))
   # without variable
   calcNodes_reduced <- model$getDependencies(target)
 }
```

# RJ sampler run code

Depending on the current value of the variable selection indicator, we use the reversible jump calculations to propose either including or excluding the variable from the model.

We fix $J(u)$ here as a distribution that doesn't change with the current state, so it will be critical to come up with a good value for the *scale* of $J(u)$ as that is simply the proposal distribution for the new coefficient in this simple RJ scheme.

```{r, rj-run, eval = FALSE}
run = function( ) {  # Reversible-jump updates:
   currentIndicator <- model[[target]]   # get current z2
   if(currentIndicator == 0) { ## propose adding z2 to model
     currentLogProb <- model$getLogProb(calcNodes_reduced)
     proposalCoef <- rnorm(1, 0, sd = scale)
     model[[target]] <<- 1
     model[[coefNode]] <<- proposalCoef
     logProbForwardProposal <- 
                dnorm(proposalCoef, 0, sd = scale, log = TRUE)
     proposalLogProb <- model$calculate(calcNodes_full)
     logAcceptProb <- proposalLogProb - 
                        currentLogProb - 
                        logProbForwardProposal
   } else {                   ## propose removing z2
     currentLogProb <- model$getLogProb(calcNodes_full)
     currentCoef <- model[[coefNode]]      # get current beta2
     logProbReverseProposal <- 
                 dnorm(currentCoef, 0, sd = scale, log = TRUE)   
     model[[target]] <<- 0      # put proposal values in model
     model[[coefNode]] <<- 0
     model$calculate(calcNodes_full) # calculate proposal log probabilities
     logAcceptProb <- model$getLogProb(calcNodes_reduced) -
                        currentLogProb + 
                        logProbReverseProposal
   }
}   
```

# The full RJ sampler

Putting it all together, here is the full code for our RJ variable selection sampler.

```{r, full-rj}
RJ_var_sel <- nimbleFunction(
 contains = sampler_BASE,
 setup = function( model, mvSaved, target, control ) {
   # target = 'indic'
   coefNode <- control$coef   # 'delta[2]'
   scale    <- control$scale  # could be made adaptive)
   # with variable
   calcNodes_full <- model$getDependencies(c(coefNode, target))
   # without variable (don't calculate prior for coefNode since not in model)
   calcNodes_reduced <- model$getDependencies(target)
 },
run = function( ) {  # Reversible-jump updates:
   currentIndicator <- model[[target]]   # get current z2
   if(currentIndicator == 0) { ## propose adding z2 to model
     currentLogProb <- model$getLogProb(calcNodes_reduced)
     proposalCoef <- rnorm(1, 0, sd = scale)
     model[[target]] <<- 1
     model[[coefNode]] <<- proposalCoef
     logProbForwardProposal <- 
                dnorm(proposalCoef, 0, sd = scale, log = TRUE)
     proposalLogProb <- model$calculate(calcNodes_full)
     logAcceptProb <- proposalLogProb - 
                        currentLogProb - 
                        logProbForwardProposal
   } else {                   ## propose removing z2
     currentLogProb <- model$getLogProb(calcNodes_full)
     currentCoef <- model[[coefNode]]      # get current beta2
     logProbReverseProposal <- 
                 dnorm(currentCoef, 0, sd = scale, log = TRUE)   
     model[[target]] <<- 0      # put proposal values in model
     model[[coefNode]] <<- 0
     model$calculate(calcNodes_full) # calculate proposal log probabilities
     logAcceptProb <- model$getLogProb(calcNodes_reduced) -
                        currentLogProb + 
                        logProbReverseProposal
    }
    accept <- decide(logAcceptProb)
    if(accept) {
      copy(from = model, to = mvSaved, row = 1, 
         nodes = calcNodes_full, logProb = TRUE)
    } else {
      copy(from = mvSaved, to = model, row = 1, 
         nodes = calcNodes_full, logProb = TRUE)
    }
 },
 methods = list(reset = function() {
 })
) 

```

# Final modification: Turning off sampling when the covariate is not in the model

Here we'll create a sampler that simply wraps around our usual slice sampler.
Note that in setup code we can create instances of other nimbleFunctions and
then use these in the run code.

```{r, nonzero-sampler}
slice_var_sel_wrapper <- nimbleFunction(
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control) {
        regular_slice_sampler <- sampler_slice(model, mvSaved, target = target, 
                                         control = control$sliceControl)
        indicatorNode <- control$indicator
    },
    run = function() {
        if(model[[indicatorNode]] == 1) regular_slice_sampler$run()
    },
    methods = list(
        reset = function() {regular_slice_sampler$reset()}
    ))
```

```{r, scopefix, echo=FALSE, fig.cap='', fig.width=5, fig.height=5}
# not clear why RW_reflect() not being put into global
# if this isn't done, configureMCMC fails to find sampler_RW_reflect in knitr
assign('slice_var_sel_wrapper', slice_var_sel_wrapper, .GlobalEnv)
assign('RJ_var_sel', RJ_var_sel, .GlobalEnv)
```

# Running the MCMC

Now let's run the MCMC. We'll need to manually assign our special samplers for the RJ aspect of things. 
      
```{r, aft-mcmc}
nSub = 15
constants = list(b0 = -10, B0 = 10, a0 = 0.1, A0 = 10, p = 2, n = n,
                 c = cens_time, Z = cbind(logBUN, HGB), nSub = nSub)
data = list(is_cens = as.numeric(alive), x = time)
xInit <- rep(NA, n)
xInit[alive] <- cens_time[alive] + 10
inits = list(alpha = 1, delta = c(0, 0), conc = 1,
             etaTilde = runif(nSub, constants$b0, constants$B0),
             xi = sample(1:3, n, replace = TRUE), x = xInit, indic = 1)
model <- nimbleModel(codeAFTsel, constants = constants, data = data, inits = inits)
cmodel = compileNimble(model)
conf <- configureMCMC(model, thin = 10, monitors = c('alpha', 'delta', 'indic'))
conf$removeSamplers(c('alpha', 'delta', 'etaTilde', 'indic'))
conf$addSampler('alpha','slice')
conf$addSampler('delta[1]','slice')
for(node in model$expandNodeNames('etaTilde'))
    conf$addSampler(node,'slice')

## special changes for variable selection:
conf$addSampler('delta[2]','slice_var_sel_wrapper', control = list(indicator = 'indic'))
conf$addSampler('indic', 'RJ_var_sel', control = list(coef = 'delta[2]', scale = 0.5))

mcmc <- buildMCMC(conf)
cmcmc <- compileNimble(mcmc, project = model)
resultsAFT <- runMCMC(cmcmc, niter = 21000, nburnin = 1000) 
```

Let's look at the MCMC behavior of the indicator and of the coefficient of interest.

```{r, results, fig.width=8, fig.height=5, fig.cap=''}
par(mfrow = c(1,2))
ts.plot(resultsAFT[ , 'indic'], xlab = 'iteration', ylab = 'z2 indicator',
                    main = 'z2 indicator')
ts.plot(resultsAFT[ , 'delta[2]'], xlab = 'iterations', ylab = 'delta[2]',
               main = 'hemoglobin coefficient')

## posterior probability of inclusion    
mean(resultsAFT[ , 'indic'])  
```

