---
title: "Running an MCMC (state-space model example)"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r loadnimble, include=FALSE}
library(nimble)
```
```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
read_chunk('chunks_ssm.R')
```

# A basic MCMC

The steps of running an MCMC are as follows:

 - configure the MCMC (can be combined with next step)
 - build the MCMC 
 - compile the MCMC
 - run the MCMC
 - extract the MCMC samples

# Build the model

We first need to build the model.

```{r, ssm-code}
```

```{r, ssm-model}
```

```{r, ssm-compile}
```

# Configuring a basic MCMC

*Configuring* the MCMC means setting up the samplers to be used for
 each node or group of nodes, which nodes will be recorded and any
 thinning.

A NIMBLE MCMC allows two separate sets of nodes to be recorded, each
with its own thinning interval.

NIMBLE provides a default configuration, but we'll see shortly how you can modify that. 

```{r, conf}
mcmcConf <- configureMCMC(ssm)
```

You can see what kind of sampler has been assigned to each node:
```{r seesamplers}
mcmcConf$printSamplers()
```

This list reveals that this model is not written well for efficient
computation.  Every time time the MCMC considers a new value for
`r[t]`, the value of `logN.est` and the probability density of
`y` must be calculated for every value of `t` until the end of the
data.  We'll reconsider that later, but it's tangential to our current
purpose of illustrating MCMC.

We'll see how to modify sampler assignments later.  

You can also inspect the monitors, which is the vector of nodes that will be
recorded in output.

```{r, seemonitors}
mcmcConf$getMonitors()
```

If we wanted to add or remove variables from monitors, we could use
`mcmcConf$addMonitors()`.

# Building, compiling and running the MCMC


```{r build-mcmc}
ssmMCMC <- buildMCMC(mcmcConf)
CssmMCMC <- compileNimble(ssmMCMC, project = ssm)
```

Now let's run the MCMC.  *Both the uncompiled and compiled versions
can be run*, but the uncompiled version will be really slow.  Its
purpose is to allow debugging of algorithms from R.  (They will usually
create exactly the same samples if you set the random number seed to
the same value before each.)

Let's run each for 10 iterations and time them:
```{r run-mcmc, warning=FALSE}
set.seed(0)
print(system.time(ssmMCMC$run(10)))  # uncompiled version
set.seed(0)
print(system.time(CssmMCMC$run(10)))   # compiled version
```

In this case the uncompiled MCMC generates a set of warnings
(suppressed in these slides).   These arise from the
Metropolis-Hastings algorithm proposing negative values for standard
deviations, which create NaNs in `dnorm` calculations.  These are not
a problem and could be removed by using a reflected sampler, a sampler
on a log scale, or other customization.

But now let's run the compiled version for much longer:
```{r run-mcmc-longer}
print(system.time(CssmMCMC$run(10000)))
```

# Extracting MCMC output

The samples are saved inside of the `CssmMCMC` object and must be
extracted and converted to a matrix.

```{r Rmcmc}
samples <- as.matrix(CssmMCMC$mvSamples)
```

Now let's look at a few parameters

```{r output-mcmc, fig.height=4, fig.width=9, fig.cap=""}
tsplot <- function(x, ...) plot(seq_along(x), x, type = 'l', ...)

par(mfrow = c(1, 3), mai = c(.6, .5, .1, .2))
tsplot(samples[ , 'mean.r'], xlab = 'iteration',
     ylab = 'mean.r', main = 'mean.r')
tsplot(samples[ , 'sigma.proc'], xlab = 'iteration',
     ylab = 'sigma.proc', main = 'sigma.proc')
tsplot(samples[ , 'sigma.obs'], xlab = 'iteration',
     ylab = 'sigma.obs', main = 'sigma.obs')
```

### Using CODA

NIMBLE does not provide any MCMC diagnostics, but one can easily use CODA or other R packages with the MCMC output from a NIMBLE MCMC.

```{r coda}
library(coda)
burnin <- 100
mcmc <- as.mcmc(samples[(burnin+1):nrow(samples), ])
crosscorr(mcmc[ , c('mean.r', 'sigma.proc', 'sigma.obs')])
effectiveSize(mcmc)
```

One could apply the commonly used Gelman-Rubin potential scale
reduction factor diagnostic.

NIMBLE does not automate running multiple chains, but one can easily
write a for-loop to do so.





