---
title: "Particle MCMC"
subtitle: "BYU Summer Institute on Applied Statistics workshop"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found 
library(nimble)
read_chunk('chunks_stoch_vol.R')
```

# Particle MCMC

Note that at each step, one can get a Monte Carlo estimate of $p(y_t|y_{1:t-1}, \theta)$, so one can multiply to estimate $p(y_{1:T}|\theta)$.

Recall that for MCMC,

   - High-dimensional latent process values in non-conjugate models often result in bad mixing.
   - Ideally, we'd like to integrate over $x_{1:T}$ and do MCMC only on hyperparameters, $\theta$.
   - SMC algorithms allow us to estimate the marginal likelihood so could be embedded within MCMC for the hyperparameters.

# Particle MCMC in NIMBLE

NIMBLE provides scalar and block random-walk Metropolis Hastings based on this approach: "Particle Marginal Metropolis Hastings".

Simply specify 'RW_PF' or 'RW_PF_block' in *addSampler*, indicating the $x_{1:T}$ nodes as part of the control argument.

We'll look directly at the PMCMC code in [`PMCMC_samplers.R`](PMCMC_samplers.R), which is simply the PMCMC samplers extracted from *MCMC_samplers.R* file in the nimble R package.

The setup code creates a filtering algorithm, and then the run code runs it under the proposed hyperparameter values and uses the likelihood approximation in the Metropolis-Hastings acceptance calculation.


# Stochastic volatility example revisited

```{r, sv-code}
```

```{r, sv-model}
```

# Stochastic volatility, particle MCMC

```{r, sv-pmcmc}
stochVolConf <- configureMCMC(stochVolModel, nodes = NULL,
    monitors = c('beta', 'phi', 'sigma' , 'x'))
stochVolConf$addSampler(target = c('beta', 'phiStar', 'sigma' , 'x0'),
                               type = 'RW_PF_block', control = list(propCov = .1 * diag(4),
                               pfType = 'bootstrap', pfControl = list(thresh = 1),
                               adaptive = TRUE, pfNparticles = 200,
                               latents = 'x', pfResample = TRUE))
                               
stochVolMCMC <- buildMCMC(stochVolConf)
cMCMC <- compileNimble(stochVolMCMC, project = stochVolModel, resetFunctions = TRUE)
samples <- runMCMC(cMCMC, niter = 5000)
```

# Stochastic volatility, particle MCMC results

It looks like the MCMC worked somewhat well.

```{r, sv-pmcmc-results, fig.width=12, fig.height=8, fig.cap=''}
par(mfrow = c(2, 3))
hist(samples[ , 'beta'])
hist(samples[ , 'phi'])
hist(samples[ , 'sigma'])
ts.plot(samples[ , 'beta'])
ts.plot(samples[ , 'phi'])
ts.plot(samples[ , 'sigma'])
```
