---
title: "Some inspirations for how you could you use NIMBLE"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
library(methods)
```
```{r loadnimble, include=FALSE}
library(nimble)
```

# What are some ways you can use NIMBLE?

   - Run your own analyses, of course.
   - Combine methods in new ways: e.g. Particle MCMC = Particle filter + MCMC.
   - Write and share new MCMC samplers.
   - Write and share new algorithms.
   - Make a package that uses NIMBLE as an engine.

# Example: Providing efficient MCMCs for a class of models

   - Say you have good MCMC strategies for state-space models
   - You could write a function to automate the custom configuration:
       + e.g., `ssmMCMC <- configureStateSpaceMCMC(ssm)`
   - You could write a package that provides the samplers and configuration function to end users.

# Example: Providing high-level access to a class of models and/or methods
   - Say you want to provide an R-style single-line call such as `lm` or `glm` for a complicated class of models.
   - You can convert that line into inputs for one or more pre-written BUGS models and then use any algorithms on it.
       + Model-definition if-then-else could be especially useful.
   - The idea is to raise the level of abstraction one level above BUGS code.

# Example: You have an algorithm that could be used for many kinds of models

  - Maybe you want to automate Bayesian cross-validation.
      + Warning: how to hold out data in a hierarchical setting is not trivial.
  - You can program this in a way that adapts to any `nimbleModel` by querying each model  about its nodes and their relationships.
  - Then you can provide the *model-generic* algorithm as a package.
