---
title: "Using our understanding of graphs to re-write the state-space model example"
subtitle: "NIMBLE training materials module"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r loadnimble, include=FALSE}
library(nimble)
```
```{r chunksetup, include=FALSE} 
# Following code is only needed for slide generation, not for using R code separately.
library(methods)
read_chunk('chunks_ssm.R')
```
# Outline

In this module, we'll apply our understanding of the computations implied by the graphical structure of a model.

The state-space model as written previously creates a computational burden that increases non-linearly with the length of the data set.

# The running state-space model example

```{r, ssm-code}
```

```{r, ssm-model}
```

# Re-writing the model

In this model, if an algorithm (such as MCMC) wants to change `r[t]` and update the model correspondingly, then it needs to recalculate `logN.est[t+1]`, `logN.est[t+2]`, `logN.est[t+3]` etc.  It also needs to recalculate the log probability density for `y[t+1]`, `y[t+2]`, `y[t+3]`, etc.

We can re-write the model like this:

```{r}
ssmCodeAlt <- nimbleCode({
# Priors and constraints
logN.est[1] ~ dnorm(5.6, 0.01)       # Prior for initial population size
mean.r ~ dnorm(1, 0.001)             # Prior for mean growth rate
sigma.proc ~ dunif(0, 1)             # Prior for sd of state process
sigma.obs ~ dunif(0, 1)              # Prior for sd of observation process

# Likelihood
# State process
for (t in 1:(T-1)){
   logN.est[t+1] ~ dnorm(logN.est[t] + mean.r, sd = sigma.proc)
   }

# Observation process
for (t in 1:T) {
   y[t] ~ dnorm(logN.est[t], sd = sigma.obs)
   }

# Population sizes on real scale
for (t in 1:T) {
   N.est[t] <- exp(logN.est[t])
   }
})
```

# Compare MCMCs

We'll use `compareMCMCs` and related functions to compare MCMC efficiency for the original and alternate way to write the model. 

```{r}
set.seed(1)
ourInits <- inits()
mcmcResult <- compareMCMCs(
     list(ssm = list(code = ssmCode, inits = ourInits, data = bugs.data)),
     niter = 50000, burnin = 5000, MCMCs = 'nimble', summary = TRUE)

mcmcResultAlt <- compareMCMCs(
     list(ssm = list(code = ssmCodeAlt, inits = ourInits, data = bugs.data)),
     niter = 50000, burnin = 5000, MCMCs = 'nimble', summary = TRUE)

mcmcResultAlt[[1]] <- rename_MCMC_comparison_method('nimble', 'nimble(alt)',
                      mcmcResultAlt[[1]])

mcmcBoth <- combine_MCMC_comparison_results(mcmcResult[[1]], mcmcResultAlt[[1]])

make_MCMC_comparison_pages(mcmcBoth, dir = 'ssm_MCMC_comparison')
```

The results are [here](ssm_MCMC_comparison/MCMCresults.html)

It does look like the alternative version yields faster MCMC.
